---
title: |
  | \huge A Brief Introduction to
  | \huge Programming with R
subtitle: "MEcon/MIQEF Introductory Week"
author: "Erik Senn & Jeremia Stalder"
date: "September 5, 2025"
format:
  beamer:
    theme: default
    colortheme: default
    fonttheme: default
    navigation: horizontal
    section-titles: true
    toc: false
    slide-level: 3
    fig-width: 5
    fig-height: 3
    fig-align: center
    df-print: kable
    fontsize: 10pt
    aspectratio: 169
    keep-tex: false
    latex-engine: xelatex
    include-in-header:
      text: |
        \usepackage{fontspec}
        \usepackage{tikz}
        \usetikzlibrary{shapes,arrows,positioning}

        % Colors
        \definecolor{mDarkBlue}{HTML}{23373b}
        \definecolor{mLightBlue}{HTML}{0066CC}
        \definecolor{mAccent}{HTML}{EB811B}
        \definecolor{mGray}{HTML}{4A4A4A}
        \definecolor{mLightGray}{HTML}{F5F5F5}
        \definecolor{mBackground}{HTML}{FAFAFA}
        \definecolor{mGreen}{HTML}{14B03D}
        \definecolor{mPurple}{HTML}{6B3AA0}

        % Fonts (fallbacks)
        \IfFontExistsTF{Fira Sans}{\setsansfont{Fira Sans}}{%
          \IfFontExistsTF{Helvetica Neue}{\setsansfont{Helvetica Neue}}{\setsansfont{Arial}}}

        % Body text size
        \setbeamerfont{normal text}{size=\normalsize}
        % All slide titles (##) as \huge
        \setbeamerfont{frametitle}{size=\huge,series=\bfseries}
        \setbeamerfont{section title}{size=\huge,series=\bfseries}
        \setbeamerfont{subsection title}{size=\huge,series=\bfseries}

        % Beamer colors
        \setbeamercolor{normal text}{fg=mDarkBlue}
        \setbeamercolor{alerted text}{fg=mAccent}
        \setbeamercolor{example text}{fg=mLightBlue}
        \setbeamercolor{structure}{fg=mDarkBlue}
        \setbeamercolor{title}{fg=white,bg=mDarkBlue}
        \setbeamercolor{frametitle}{fg=mDarkBlue,bg=mBackground}
        \setbeamercolor{section in toc}{fg=mDarkBlue}
        \setbeamercolor{subsection in toc}{fg=mGray}
        \setbeamertemplate{navigation symbols}{}

        % Frame title with rule
        \setbeamertemplate{frametitle}{
          \vspace{0.4em}\insertframetitle\vspace{0em}
          \textcolor{mAccent}{\rule{\textwidth}{2pt}}
        }

        % Section page
        \AtBeginSection[]{
          \begin{frame}[plain]
            \vfill\centering
            \begin{beamercolorbox}[sep=12pt,center,shadow=false,rounded=true]{title}
              {\Huge\bfseries\insertsectionhead\par}
            \end{beamercolorbox}
            \vfill
          \end{frame}
        }

        % Footline
        \setbeamertemplate{footline}{
          \hbox{%
          \begin{beamercolorbox}[wd=.5\paperwidth,ht=2.25ex,dp=1ex,left]{author in head/foot}%
            \hspace{2ex}\usebeamerfont{author in head/foot}\insertshortauthor
          \end{beamercolorbox}%
          \begin{beamercolorbox}[wd=.5\paperwidth,ht=2.25ex,dp=1ex,right]{date in head/foot}%
            \usebeamerfont{date in head/foot}\insertframenumber{} / \inserttotalframenumber\hspace{2ex}
          \end{beamercolorbox}}%
        }

        % Lists
        \setbeamertemplate{itemize item}{\textcolor{mLightBlue}{$\bullet$}}
        \setbeamertemplate{itemize subitem}{\textcolor{mGray}{$\circ$}}
        \setbeamertemplate{itemize subsubitem}{\textcolor{mGray}{$-$}}
        \setbeamertemplate{enumerate item}{\textcolor{mLightBlue}{\arabic{enumi}.}}
        \setbeamertemplate{enumerate subitem}{\textcolor{mGray}{\alph{enumii}.}}
        \setlength{\itemsep}{0.5em}
        \setlength{\parskip}{0.3em}

        % Blocks
        \setbeamercolor{block title}{fg=white,bg=mLightBlue}
        \setbeamercolor{block body}{fg=mDarkBlue,bg=mLightGray}
        \setbeamercolor{block title alerted}{fg=white,bg=mAccent}
        \setbeamercolor{block body alerted}{fg=mDarkBlue,bg=mAccent!10}

        % Custom text macros (used in slides)
        \newcommand{\keyword}[1]{\textcolor{mLightBlue}{\textbf{#1}}}
        \newcommand{\accenttext}[1]{\textcolor{mAccent}{\textbf{#1}}}
        \newcommand{\highlight}[1]{\textcolor{mGreen}{\textbf{#1}}}
        \newcommand{\important}[1]{\textcolor{mPurple}{\textbf{#1}}}

        % ---- Code blocks: normalsize + extra vertical space via hooks ----
        \usepackage{etoolbox}

        % Default size
        \newcommand{\CodeSize}{\normalsize}
        \newcommand{\SetCodeSize}[1]{\gdef\CodeSize{#1}}
        
        % Hooks for Pandoc's Highlighting environment
        \AtBeginEnvironment{Highlighting}{\vspace{2pt}\CodeSize}
        \AfterEndEnvironment{Highlighting}{\vspace{2pt}\SetCodeSize{\normalsize}}
        
        % Also adjust font used inside
        \makeatletter
        \def\verbatim@font{\CodeSize\ttfamily}
        \makeatother

classoption:
  - compress
  - fragile
execute:
  echo: true
  warning: false
  message: false
  comment: ""
  R.options:
    width: 50
---

```{r setup, include=FALSE}
library(ggplot2)
library(readxl)
library(haven)
library(stargazer)
library(knitr)
library(tidyverse)

# Create a modern theme for plots
theme_modern <- function() {
  theme_minimal(base_size = 10) +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.grid.major = element_line(color = "#E0E0E0", size = 0.3),
      panel.grid.minor = element_blank(),
      text = element_text(family = "sans", color = "#23373b"),
      axis.text = element_text(color = "#4A4A4A"),
      axis.title = element_text(face = "bold", size = 11),
      plot.title = element_text(face = "bold", size = 14, hjust = 0),
      plot.subtitle = element_text(size = 12, hjust = 0, color = "#4A4A4A"),
      legend.position = "bottom",
      legend.title = element_text(face = "bold"),
      strip.text = element_text(face = "bold")
    )
}

theme_set(theme_modern())
set.seed(123)

# Modern color palette
modern_colors <- c("#0066CC", "#EB811B", "#14B03D", "#CC0033", "#6B3AA0", "#F7B500")
scale_color_modern <- function(...) scale_color_manual(values = modern_colors, ...)
scale_fill_modern <- function(...) scale_fill_manual(values = modern_colors, ...)

options(
  width = 50,
  scipen = 999,
  digits = 3
)

opts_chunk$set(
  tidy = FALSE,
  size = "normalsize"
)
```

# Part I: Background / Tools {.section}

### Schedule

\begin{columns}[T]
\begin{column}{0.48\textwidth}
\textbf{Morning Sessions}

\textcolor{mLightBlue}{09:15 - 10:00} \\
Introduction, Background, Tools

\textcolor{mLightBlue}{10:00 - 10:15} \\
Break, support with installations

\textcolor{mLightBlue}{10:15 - 11:00} \\
Exercises, First steps with R

\textcolor{mLightBlue}{11:00 - 11:15} \\
Break, Q\&A

\textcolor{mLightBlue}{11:15 - 12:00} \\
First steps with R, Concepts
\end{column}

\begin{column}{0.48\textwidth}
\textbf{Afternoon Sessions}

\textcolor{mLightBlue}{12:00 - 13:15} \\
Lunch (individually)

\textcolor{mLightBlue}{13:15 - 14:00} \\
Exercises

\textcolor{mLightBlue}{14:00 - 14:15} \\
Break, Q\&A

\textcolor{mLightBlue}{14:15 - 15:00} \\
Working with Data
\end{column}
\end{columns}

### Welcome

\begin{enumerate}
\tightlist
\item \keyword{Fire up} your notebooks!

\vspace{0.3em}

\item \keyword{Download} (or clone) the course materials
  \begin{itemize}
  \item \href{https://github.com/ErikSenn/intro_to_r_2025/}{\textcolor{mLightBlue}{GitHub Repository}}
  \item Course slides available online
  \end{itemize}
\end{enumerate}

## Why learn to program (now)?

### Background: Technological change

:::::: {.columns}
::: {.column width="60%"}
\vspace{1em}
\Large
\begin{itemize}
\item Computers have become \keyword{omnipresent}

\vspace{0.3em}

\item Data is one of the world's most \accenttext{valuable resources}

\vspace{0.3em}

\item \highlight{AI and machine learning} are reshaping every business and industry
\end{itemize}
:::

::: {.column width="40%"}

:::
::::::

### Perspective: Data Science and Economics

\begin{overprint}
  \only<1>{\includegraphics[width=.9\textwidth]{img/Slide7.jpg}}
  \only<2>{\includegraphics[width=.9\textwidth]{img/Slide6.jpg}}
  \only<3>{\includegraphics[width=.9\textwidth]{img/Slide5.jpg}}
  \only<4>{\includegraphics[width=.9\textwidth]{img/Slide4.jpg}}
  \only<5>{\includegraphics[width=.9\textwidth]{img/Slide3.jpg}}
  \only<6>{\includegraphics[width=.9\textwidth]{img/Slide2.jpg}}
  \only<7>{\includegraphics[width=.9\textwidth]{img/Slide1.jpg}}
\end{overprint}

## Why learn to program if AI assis


## Why R?

### A data language

:::::: {.columns}
::: {.column width="50%"}
\vspace{1em}
- Widely used in \keyword{data science} jobs

\vspace{0.5em}

- Particularly adapted to program with \accenttext{data}

\vspace{0.5em}

- Originally designed for \highlight{statistical analysis}

\vspace{0.5em}

- Competing with Python as the \important{top data science language}
:::

::: {.column width="50%"}
\includegraphics[width=.5\textwidth]{img/R_logo.png}
\vspace{1em}
\includegraphics[width=.5\textwidth]{img/Python_logo.png}
:::
::::::

### A high-level language

:::::: {.incremental}
- \keyword{Relatively easy} to learn

\vspace{0.3em}

- Extensive \accenttext{free resources}:
  - \href{https://www.datacamp.com/courses/free-introduction-to-r}{\textcolor{mLightBlue}{DataCamp: Introduction to R}}
  - \href{https://www.rstudio.com/resources/cheatsheets/}{\textcolor{mLightBlue}{RStudio Cheatsheets}}
  - \href{https://stackoverflow.com/}{\textcolor{mLightBlue}{Stack Overflow}}
::::::


## The Tools

### R

\keyword{R} is the \accenttext{programming language.}

You can download R from: [https://cran.r-project.org/](https://cran.r-project.org/)

### RStudio Desktop

\keyword{RStudio} is the \accenttext{integrated development environment (IDE).}

You can download RStudio Desktop from: [https://posit.co/download/rstudio-desktop/](https://posit.co/download/rstudio-desktop/)

### RStudio Cloud

\keyword{RStudio Cloud} lets you use RStudio \highlight{without a local installation}

You can use RStudio (Posit) Cloud by registering here: [https://posit.cloud/](https://posit.cloud/)


## An RStudio Tour

<!--

\frametitle{RStudio}

\vspace{-15pt}
\begin{itemize}
	\item the most popular IDE to work in R
	\item IDE (integrated development environment) = software application for software development integrating several developer tools into one user interface
	\item opening R Studio opens an R Session with:
	\begin{itemize}
		\item \textbf{Console Window} = here the commands are typed and results are returned
		\item \textbf{Environment / History} = shows all objects created in the current R session (Environment tab) and all commands used (History tab)
		\item \textbf{Files / Plots / Packages / Help} = primarily used to see the files system (Files tab), created plots (Plots tab), install or check available packages (Packages tab) and documentation (Help tab) 
	\end{itemize}
\end{itemize}
}

\frame{
	\begin{center}
		\includegraphics[width=\textwidth]{RstudioGUI.png}
	\end{center}

%	\textbf{Task:} Type 2+2 into the console and press Enter.

}

\Section{R Scripts}

\frame{
	
	\frametitle{R Scripts}
	\vspace{-15pt}
	\begin{itemize}
		\item R code can be written directly into the R Console or it can be typed as a simple text file (R Script) that is later run in the console
		\item advantage of R scripts: code is saved in a file that can be re-run, augmented and shared at any time.
	\end{itemize}
	
}




\frame{
	
	\begin{center}
		\textbf{Task 1:}
		\vspace{10pt}
		\begin{itemize}
			\centering
			\item Click on \textbf{File - New File - R Script} \\
			\item Save the new R script (File - Save or Ctrl + S or CMD + S on MacOS) into the folder ``01-Basics'' under a name ``01-Basics.R''
		\end{itemize}
	\end{center}
}

\frame{
	
	\begin{center}
		\includegraphics[width=\textwidth]{RscriptWin.png}
	\end{center}
	
}

\frame{
	
	\frametitle{Writing and Running Scripts}
	
	\vspace{-15pt}
	\begin{itemize}
		\item any text after \# is interpreted as a comment
		\item any other text is interpreted as active code
		\item to run the whole script press Ctrl + Shift + Enter (CMD + Shift + Enter on MacOS)
		\item to run only a part of the script, select all lines that should run and press Ctrl + Enter (CMD + Enter on MacOS)
	\end{itemize}
	
}

\Section{Working Directory}

\frame{
	
	\frametitle{Working Directory}
	
	\vspace{-15pt}
	\begin{itemize}
		\item R will save your reults into the current working directory
		\item to find out what the current working directory is, run {\color{blue} \texttt{getwd()}} in the console
		\item changing the working directory can be done via
		\begin{itemize}
			\item Session > Set Working Directory > Choose Directory, click to the new directory and confirm by clicking on Open
			\item or use function {\color{blue} \texttt{setwd()}} with the path leading to the new directory
			\item[] E.g. {\color{blue} \texttt{setwd(``c:/user/R/Tutorial1'')}}
		\end{itemize}
	\end{itemize}
	
}

\Section{R Help}

\frame{
	\frametitle{R help}
	
	\vspace{-15pt}
	\begin{itemize}
		\item if the name of the function is known, type \texttt{?} and the name of the function 
		\item[] E.g. {\color{blue} \texttt{?getwd}}
		\item this will open the help for the function
		\item if function is in a package, clicking in the Packages menu on the package and the function leads to its documentation
		\item otherwise, Google is a good friend
	\end{itemize}
	
}

## R-projects / relation to paths/working directory

## Possibly: reference to useful features - git, data import visual interface, git, automatic AI-assistance.


-->


## Exercises

### Exercise A: Setting up a Working Environment

\begin{enumerate}
\tightlist
\item Open \keyword{RStudio} and navigate to your desired working folder

\vspace{0.3em}

\item Create a new folder called \accenttext{r\_course}

\vspace{0.3em}

\item Set it as your \keyword{working directory}

\vspace{0.3em}

\item Create subfolders: \highlight{data} and \highlight{code}

\vspace{0.3em}

\item Create a new \important{R Project} in the \texttt{r\_course} folder
\end{enumerate}
\normalsize

### Exercise B: R Scripts

1. In the R console, type:

```{r, eval=FALSE}
print("Hello world")
```

2. Create a new \keyword{R script}: `File > New File > R Script`

3. Type the same code in the script and \accenttext{run it}

\normalsize

# Part II: First Steps and Basic Concepts {.section}

## First Steps in R

### Variables and Vectors

```{r}
a <- 2  
a       
```
\normalsize

### Working with Vectors

R easily allows to work with \keyword{vectors} of data!

```{r}
# Instantiate an integer vector 'a'
a <- c(10, 22, 33, 22, 40)
# Give names to the vector's elements
names(a) <- c("Andy", "Betty", "Claire", 
              "Daniel", "Eva")
a  # Display the vector
```
\normalsize

### Indexing

We can access \keyword{single} (or \accenttext{multiple}) elements of a vector:

```{r}
a[3]     # Access the 3rd element
a[3:5]   # Access the 3rd to 5th elements
a["Claire"]  # Access by name
```
\normalsize

### Inspecting variables

```{r}
class(a)  # Display the class

str(a)    # Display the structure
```
\normalsize

### Math Operators

\keyword{Basic operators:}

- $+$ : `+`
- $-$ : `-`  
- $\times$ : `*`
- $\div$ : `/`

\accenttext{More operators:}

- $a^n$ : `a^n`
- $\sqrt{a}$ : `sqrt(a)`
- $\ln a$ : `log(a)`
- $e^n$ : `exp(n)`

## Basic Programming Concepts

### Loops

- \keyword{Repeatedly execute} a sequence of commands

\vspace{0.3em}

- For a \accenttext{known} or \accenttext{unknown} number of iterations
  - \highlight{for-loop}: number of iterations typically known
  - \highlight{while-loop}: iterate until a condition is met

### for-loops in R

```{r}
n_iter <- 5  # Define number of iterations
# Specify the loop
for (i in 1:n_iter) {
    print(i)  # Print the number 'i'
}
```
\normalsize

### for-loops: Summing numbers

```{r}
numbers <- c(72, 42, 150, 13, 36, 19)
total_sum <- 0  # Initialize sum

# Specify the loop
for (n in numbers) {
    total_sum <- total_sum + n
}
total_sum
```
\normalsize

### Nested for-loops

```{r, eval=FALSE}
n_iter_inner <- 100
n_iter_outer <- 500

# Start outer loop
for (i in 1:n_iter_outer) {
    # Code for outer loop
    
    # Start inner loop
    for (j in 1:n_iter_inner) {
        # Code for inner loop
    }
}
```
\normalsize

How many iterations does this combination amount to?

<!--
### while-loops in R

```{r}
# Instantiate starting value
total <- 0

# Loop while the condition is TRUE
while (total <= 10) {
    total <- total + 3.14
}
total
```
\normalsize
-->

### Booleans and Logical Statements

\footnotesize
```{r}
2 + 2 == 4  # Is 2+2 equal to 4?
3 + 3 == 7  # Is 3+3 equal to 7?
4 != 7      # Is 4 not equal to 7?
```
\normalsize

### Control Flow with Booleans

\footnotesize
```{r}
condition <- TRUE

if (condition) {
    print("The condition is true!")
} else {
    print("The condition is false!")
}
```
\normalsize

### R Functions

- Functions take \keyword{parameter values} as input, process these values, and \accenttext{return} results

\vspace{0.3em}

- Many functions are \highlight{provided with R}

\vspace{0.3em}

- Additional functions via \important{packages}

\footnotesize
```{r}
numbers <- c(13, 25, 39, 881)
mean(numbers)    # Compute the mean
sd(numbers)      # Standard deviation
median(numbers)  # Median
```
\normalsize

### Creating Custom Functions

\footnotesize
```{r}
# Define custom mean function
my_mean <- function(x) {
    x_bar <- sum(x) / length(x)
    return(x_bar)
}

# Test the function
my_mean(numbers)
mean(numbers)  # Compare with built-in
```
\normalsize

## Data Structures and Indices

### Vectors and Lists

\footnotesize
```{r}
# Integer vector
integer_vector <- 9:20
integer_vector[2]    # Second element
integer_vector[2:5]  # Second to fifth

# String vector
string_vector <- c("a", "b", "c")
string_vector[-3]    # All except third
```
\normalsize

### Lists

Lists can contain \keyword{different types} of elements:

\footnotesize
```{r}
# Create a list
my_list <- list(
    numbers = integer_vector,
    letters = string_vector,
    condition = TRUE
)
str(my_list)
```
\normalsize

### Accessing List Elements

\footnotesize
```{r}
# Access by name
my_list$numbers[1:3]
my_list[["letters"]]

# Access by index
my_list[[1]][1:3]
```
\normalsize

### Matrices

\footnotesize
```{r}
# Create a matrix
my_matrix <- matrix(integer_vector, nrow = 4)
my_matrix

my_matrix[2,]      # Second row
my_matrix[, 1:2]   # First two columns
```
\normalsize

### Data Frames

\footnotesize
```{r}
# Create a dataframe
my_df <- data.frame(
    Name = c("Alice", "Betty", "Claire"),
    Age = c(20, 30, 45)
)

my_df
my_df$Age         # Access column
my_df[2, ]        # Second row
```
\normalsize

## Exercises

### Exercise A: Write a Sum Function

Write a function that takes a \keyword{numeric vector} as input and returns the \accenttext{sum} of the vector's elements.

```{r, eval=FALSE}
my_sum <- function(x) {
    # Your code here
}
```

Test by comparing with built-in `sum()` function.

### Exercise B: Robustness and Warnings

Test your `my_sum()` function with:

```{r, eval=FALSE}
numbers2 <- c("1", "2", "3")
```

Add \keyword{error checking} to make the function robust.

### Exercise C: Standard Deviation Function

Implement a function to compute the \accenttext{standard deviation}:

$$\text{SD} = \sqrt{\frac{1}{N-1}\sum_{i=1}^N (x_i - \bar{x})^2}$$

### Exercise D: Standard Error Function

Building on Exercise C, implement:

$$\text{SE}_{\bar{x}} = \frac{\text{SD}}{\sqrt{N}}$$

### Exercise E: T-test

Implement a \keyword{one-sample t-test} function:

$$t = \frac{\bar{x} - \mu_0}{\text{SE}_{\bar{x}}}$$

### Exercise F: Fibonacci Sequence

Generate the first 30 \accenttext{Fibonacci numbers} where:

- $F_0 = 0$, $F_1 = 1$
- $F_n = F_{n-1} + F_{n-2}$ for $n > 1$

### Exercise G: Multiples of 3 and 5 (*)

Write a function that computes the sum of all \keyword{multiples of 3 or 5} up to a number $N$.

\accenttext{Hint:} Multiples of both 3 and 5 should only be added once!

### Exercise H: Prime numbers (*)

Write a function that computes the sum of all \highlight{prime numbers} up to a number $N$.

# Part III: Working with Data {.section}

## Loading/Importing Data

### Loading built-in R data sets

\footnotesize
```{r}
# Load built-in dataset
data(swiss)

# Check if loaded
class(swiss)
```
\normalsize

The `data()` function loads \keyword{built-in R datasets} into your environment.

### Inspect the data structure

\footnotesize
```{r}
# Structure of the swiss dataset
str(swiss)
```
\normalsize

### First few rows of the data

\footnotesize
```{r}
head(swiss, 4)
```
\normalsize

### Comma Separated Values (CSV)

Example of \keyword{CSV format}:

\footnotesize
```
"","Fertility","Agriculture","Examination",...
"Courtelary",80.2,17,15,...
"Delemont",83.1,45.1,6,...
```
\normalsize

### Importing data from different sources

\footnotesize
```{r, eval=FALSE}
# CSV files
swiss_csv <- read.csv("./data/swiss.csv")

# Excel files (requires readxl)
library(readxl)
swiss_excel <- read_excel("./data/swiss.xlsx")

# SPSS files (requires haven)
library(haven)
swiss_spss <- read_spss("./data/swiss.sav")
```
\normalsize

## Introduction to the Tidyverse

### What is the Tidyverse?

To install and load:

```{r, eval=FALSE}
install.packages("tidyverse")
library(tidyverse)
```

### The Pipe Operator `|>`

Transform nested functions into \keyword{readable pipelines}:

```{r, size='footnotesize'}
# Traditional approach
head(select(swiss, Fertility, Education), 3)

# With pipe operator
swiss |> 
  select(Fertility, Education) |> 
  head(3)
```
\normalsize

### Select columns with `select()`

\footnotesize
```{r}
# Select specific columns
swiss |> 
  select(Fertility, Education, Catholic) |>
  head(3)
```
\normalsize

### Select columns - Advanced patterns

\footnotesize
```{r}
# Select by pattern
swiss |> 
  select(starts_with("E"), contains("Mort")) |>
  head(3)
```
\normalsize

### Filter rows with `filter()`

\footnotesize
```{r}
# Keep rows meeting conditions
swiss |> 
  filter(Education > 20) |>
  select(Fertility, Education, Catholic) |>
  head()
```
\normalsize

### Create new variables with `mutate()`

\footnotesize
```{r}
swiss |> 
  mutate(
    High_Education = Education > 15,
    Fert_per_100 = Fertility / 100
  ) |>
  select(Education, High_Education, 
         Fertility, Fert_per_100) |>
  head(3)
```
\normalsize

### Arrange rows with `arrange()`

\footnotesize
```{r}
swiss |> 
  arrange(desc(Education)) |>
  select(Education, Examination) |>
  head(4)
```
\normalsize

### Summarize data with `summarize()`

\footnotesize
```{r}
swiss |> 
  summarize(
    mean_edu = mean(Education),
    sd_edu = sd(Education),
    median_fert = median(Fertility),
    n = n()
  )
```
\normalsize

### Group operations with `group_by()`

\footnotesize
```{r}
# Add Religion variable
swiss <- swiss |>
  mutate(Religion = ifelse(Catholic > 50, 
                          'Catholic', 
                          'Protestant'))

# Group and summarize
swiss |> 
  group_by(Religion) |>
  summarize(
    mean_education = mean(Education),
    mean_fertility = mean(Fertility),
    count = n()
  )
```
\normalsize

### Combining multiple operations

\footnotesize
```{r}
swiss |> 
  filter(Agriculture < 50) |>
  group_by(Religion) |>
  summarize(
    avg_exam = mean(Examination),
    n = n(),
    .groups = "drop"
  ) |>
  arrange(desc(avg_exam))
```
\normalsize

### Reshape data: Wide to Long

\footnotesize
```{r}
# Original wide format
swiss_subset <- swiss |> 
  slice(1:2) |>
  select(Fertility, Agriculture)

swiss_subset
```
\normalsize

### Reshape data: `pivot_longer()`

\footnotesize
```{r}
# Convert to long format
swiss_subset |>
  pivot_longer(
    cols = everything(),
    names_to = "Metric",
    values_to = "Value"
  )
```
\normalsize

### Join data frames

\footnotesize
```{r}
# Sample data
df1 <- data.frame(
  Province = c("A", "B"),
  Population = c(5000, 6000)
)

df2 <- data.frame(
  Province = c("A", "B", "C"),
  Language = c("FR", "FR", "DE")
)

# Left join keeps all rows from df1
left_join(df1, df2, by = "Province")
```
\normalsize

### Working with dates using lubridate

\footnotesize
```{r}
library(lubridate)

# Create and parse dates
dates <- c("2023-09-08", "2023-09-15")
parsed_dates <- ymd(dates)

# Extract components
data.frame(
  date = parsed_dates,
  year = year(parsed_dates),
  month = month(parsed_dates),
  weekday = wday(parsed_dates, label = TRUE)
)
```
\normalsize

### Advanced tidyverse: map functions

\footnotesize
```{r}
# Apply a function to multiple columns
swiss |>
  select(Fertility, Agriculture, Education) |>
  map_dbl(~ mean(.x)) |>
  round(2)
```
\normalsize

### String manipulation with stringr

\footnotesize
```{r}
library(stringr)

# Example text
provinces <- c("Courtelary", "Delemont", "Franches-Mnt")

# String operations
data.frame(
  original = provinces,
  lower = str_to_lower(provinces),
  length = str_length(provinces),
  contains_e = str_detect(provinces, "e")
)
```
\normalsize

## Visualization with R (ggplot2)

### Grammar of Graphics - Layers

\center
\includegraphics[width=.7\textwidth]{img/grammar_of_graphics.png}

### ggplot2 basics

\footnotesize
```{r, eval=FALSE}
# Basic structure
ggplot(data = <DATA>, 
       mapping = aes(x = <X>, y = <Y>)) +
  <GEOM_FUNCTION>() +
  <OPTIONAL_LAYERS>
```
\normalsize

### Prepare the data

\footnotesize
```{r}
# Add Religion indicator
swiss$Religion <- factor(
  ifelse(swiss$Catholic > 50, 
         'Catholic', 'Protestant')
)

# Check the data
table(swiss$Religion)
```
\normalsize

### Building a plot: Data + Aesthetics

\footnotesize
```{r data-aes, fig.height=2}
ggplot(data = swiss, 
       aes(x = Education, y = Examination))
```
\normalsize

### Adding Geometries

\footnotesize
```{r geom, fig.height=2}
ggplot(data = swiss, 
       aes(x = Education, y = Examination)) + 
  geom_point(size = 3, alpha = 0.7, color = "#0066CC")
```

### Using Facets

\SetCodeSize{\footnotesize}
```{r facet, fig.height=2}
ggplot(data = swiss, 
       aes(x = Education, y = Examination)) + 
  geom_point(size = 3, alpha = 0.7, aes(color = Religion)) +
  scale_color_modern() +
  facet_wrap(~Religion)
```

### Adding Statistics with loess

\footnotesize
```{r stat-loess, fig.height=2}
ggplot(data = swiss, 
       aes(x = Education, y = Examination)) + 
  geom_point() +
  geom_smooth(method = 'loess') +
  facet_wrap(~Religion)
```
\normalsize

### Adding Statistics with lm

\footnotesize
```{r stat-lm, fig.height=2}
ggplot(data = swiss, 
       aes(x = Education, y = Examination)) + 
  geom_point() +
  geom_smooth(method = 'lm', se = FALSE) +
  facet_wrap(~Religion)
```
\normalsize

### Multiple Aesthetics

\footnotesize
```{r multi-aes, fig.height=2}
ggplot(data = swiss, 
       aes(x = Education, y = Examination)) + 
  geom_point(aes(color = Agriculture), size = 2) +
  geom_smooth(method = 'lm', se = FALSE) +
  facet_wrap(~Religion) +
  scale_color_gradient(low = "blue", high = "red") +
  theme(legend.position = "right")
```
\normalsize

### Change coordinates

\footnotesize
```{r coord-flip, fig.height=2}
ggplot(data = swiss, 
       aes(x = Education, y = Examination)) + 
  geom_point(aes(color = Agriculture)) +
  geom_smooth(method = 'lm', se = FALSE) +
  facet_wrap(~Religion) + 
  coord_flip()
```
\normalsize

### Customizing Themes

\footnotesize
```{r theme-custom, fig.height=2}
ggplot(data = swiss, 
       aes(x = Education, y = Examination)) + 
  geom_point(aes(color = Agriculture)) +
  geom_smooth(method = 'lm', se = FALSE) +
  facet_wrap(~Religion) + 
  theme_minimal() +
  theme(legend.position = "bottom")
```
\normalsize

### Pre-built Themes

\footnotesize
```{r theme-dark, fig.height=2}
ggplot(data = swiss, 
       aes(x = Education, y = Examination)) + 
  geom_point(aes(color = Agriculture)) +
  geom_smooth(method = 'lm', se = FALSE) +
  facet_wrap(~Religion) + 
  theme_dark()
```
\normalsize

### Save plots

\footnotesize
```{r, eval=FALSE}
# Save the last plot
ggsave("my_plot.png", 
       width = 8, height = 5, 
       dpi = 300)

# Save a specific plot
p <- ggplot(swiss, aes(Education, Examination)) + 
  geom_point()

ggsave("scatter.pdf", plot = p,
       width = 6, height = 4)
```
\normalsize

## Basic Statistics with R

### Descriptive Statistics

\footnotesize
```{r}
# Create sample data
x <- c(10, 22, 33, 22, 40)

# Basic statistics
c(mean = mean(x),
  median = median(x),
  sd = sd(x),
  min = min(x),
  max = max(x))
```
\normalsize

### T-test Example

\footnotesize
```{r}
# Generate sample data
set.seed(123)
sample <- rnorm(30, mean = 10, sd = 2)

# One-sample t-test
t_result <- t.test(sample, mu = 10)
t_result$p.value
t_result$conf.int
```
\normalsize

### Linear Regression - Setup

\footnotesize
```{r}
# Define the model formula
model1 <- Examination ~ Education

# Fit the model
fit1 <- lm(model1, data = swiss)

# View coefficients
coef(fit1)
```
\normalsize

### Simple Linear Regression - Results

\tiny
```{r}
summary(fit1)
```
\normalsize

### Multiple Linear Regression

\footnotesize
```{r}
# Fit model with multiple predictors
fit2 <- lm(Examination ~ Education + Catholic + 
           Agriculture, data = swiss)

# Model fit statistics
c(R2 = summary(fit2)$r.squared,
  Adj_R2 = summary(fit2)$adj.r.squared,
  RMSE = sigma(fit2))
```
\normalsize

### Regression Coefficients Table

\footnotesize
```{r}
# Extract coefficient information
coef_summary <- summary(fit2)$coefficients
round(coef_summary, 3)
```
\normalsize

### Other Econometric Models

\footnotesize
```{r, eval=FALSE}
# Binary outcomes - Logit
glm(y ~ x1 + x2, 
    family = binomial(link = "logit"))

# Binary outcomes - Probit  
glm(y ~ x1 + x2, 
    family = binomial(link = "probit"))

# Count data - Poisson
glm(y ~ x1 + x2, 
    family = poisson())

# Panel data - Fixed effects (requires plm)
library(plm)
plm(y ~ x1 + x2, 
    data = panel_data,
    model = "within")
```
\normalsize

### Print Regression Results - Console

\tiny
```{r, results='asis'}
# Basic regression table
stargazer(fit1, fit2, 
          type = "text",
          digits = 2,
          star.cutoffs = c(0.05, 0.01, 0.001),
          omit.stat = c("ser", "f"),
          column.sep.width = "1pt")
```
\normalsize

### Export Results to LaTeX

\footnotesize
```{r, eval=FALSE}
# For LaTeX documents
stargazer(fit1, fit2,
          type = "latex",
          title = "Regression Results",
          label = "tab:reg",
          digits = 2,
          header = FALSE,
          font.size = "small",
          column.sep.width = "1pt",
          single.row = TRUE)
```
\normalsize

### Save Regression Output

\footnotesize
```{r, eval=FALSE}
# Save as LaTeX file
stargazer(fit1, fit2,
          type = "latex",
          out = "./results/regression_table.tex",
          title = "Swiss Data Analysis",
          covariate.labels = c("Education (years)",
                              "Catholic (%)",
                              "Agriculture (%)"))

# Save as HTML
stargazer(fit1, fit2,
          type = "html",
          out = "./results/regression_table.html")
```
\normalsize

## Final Remarks

### Best Practices for R Programming

\small
1. \keyword{Organization}
   - Start scripts with `library()` calls
   - Use meaningful variable names
   - Comment your code with `#`

\vspace{0.3em}

2. \accenttext{Efficiency}
   - Use functions to avoid repetition
   - Vectorize operations when possible
   - Use the tidyverse for data manipulation

\vspace{0.3em}

3. \highlight{Reproducibility}
   - Set seeds for random operations
   - Use relative paths (`./data/`)
   - Document package versions
\normalsize

### Resources for Learning R

:::::: {.columns}
::: {.column width="50%"}
\textcolor{mLightBlue}{\textbf{Online Resources:}}
\begin{itemize}
\item \keyword{Stack Overflow}

\vspace{0.3em}

\item \accenttext{R for Data Science} book

\vspace{0.3em}

\item \highlight{RStudio cheatsheets}

\vspace{0.3em}

\item \important{CRAN documentation}
\end{itemize}
:::

::: {.column width="50%"}
\textcolor{mAccent}{\textbf{AI Assistance:}}
\begin{itemize}
\item \keyword{ChatGPT}

\vspace{0.3em}

\item \accenttext{GitHub Copilot}

\vspace{0.3em}

\item \highlight{Google Bard}
\end{itemize}
:::
::::::

### An example on AI Assistance for learning coding



### Upcoming Courses

\keyword{Next Week}

\accenttext{R Programming Course}  
Instructor: Erik Senn  
Dates: September 8-11, 2025  
Time: 9:00-12:00 and 13:00-17:00  
Location: Rosenbergstrasse 30, Room 61-152
Contents: Similar contents as today in slower pace. Automated reports using R-markdown on last day.

### Thank You!

\begin{center}
\Large
\textcolor{mLightBlue}{Thanks for joining our R workshop!}

\vspace{1em}



\vspace{1em}
\textcolor{mAccent}{\Large Happy Coding!}
\end{center}

### Survey

\begin{center}
\Large
\textcolor{mLightBlue}{Please take a moment to provide feedback on today's workshop.}

\vspace{1em}

\keyword{Your input helps us improve future courses!}
\end{center}